|

### 2023-09-06 paper.Rmd
finished revising the paper

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the availability of many numerical hypothesis test procedures designed to use residuals to assess problems with a model fit. Here we provide evidence for why this is good advice using data from a visual inference experiment. We show how conventional tests are too sensitive, which means that too often the conclusion would be that the model fit is inadequate. The experiment uses the lineup protocol which puts a residual plot in the context of null plots. This helps generate reliable and consistent reading of residual plots for better model diagnosis. It can also help in an obverse situation where a conventional test would fail to detect a problem with a model due to contaminated data. The lineup protocol also detects a range of departures from good residuals simultaneously.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  % \doublespacing
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

In addition, scatterplots of residuals plotted against the fitted values and each of the explanatory variables, can be used to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially inadequate or incorrectly specified. \textcolor{red}{Interesting features that are not directly connected to a linear model assumption can also be found from residual plots, such as the discreteness of the fitted values, outliers and influential points.} In general, one looks for noticeable departures from the model such as non-linear pattern or heteroskedasticity. A non-linear pattern would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption.

However, correctly judging whether NO pattern exists in a residual plot is a difficult task for humans. Humans will almost always see a pattern (see @kahneman2011thinking), so the question that really needs answering is whether any pattern perceived is consistent with randomness, purely sampling variability, or noise. It is especially difficult to teach this to new analysts and students [@loy2021bringing]. To address this, statistical tests are available to check for non-linear patterns [e.g. @ramsey1969tests], and heteroskedasticity  [e.g. @breusch1979simple].

\textcolor{red}{There are other types of residual plot to extract the available information and identify inappropriate assumptions, like the added variable plot (cite here) for studying the role of an external variable if it enters linearly into a model. Partial residual plots (cite here) have been suggested as computationally convenient substitutes for the added variable plots. For use in diagnostic procedures, several transformations of the ordinary residuals have been suggested to overcome partially some of their shortcomings. Studentized residuals (cite here) have null distributions that are independent of the scale parameters. These residuals are shown to be closely related to a mean shift model for outliers and to the residuals obtained when each case in turn is left out of the data. Alternatively. the residuals can be transformed to have a selected covariance structure: The usual suggestion is to obtain residuals with uncorrelated elements.}

There is sustainable literature describing diagnostic procedures for linear models, e.g. @draper1998applied, @montgomery1982introduction, @belsley1980regression, @cook1999applied and @cook1982residuals. Interestingly, despite the abundance of conventional tests, ALL of these writings advise that plotting residuals is an essential tool for diagnosing regression model problems:

@@

\noindent The common wisdom of experts is that plotting the residuals is indispensable for diagnosing model fits. The lack of empirical evidence for the ubiquitous advice is *curious*. The research study discussed in this paper fills this gap.

@@

In hypothesis testing, a $p$-value is defined as the probability of observing test results at least as extreme as the observed result assuming $H_0$ is true. Conventional hypothesis tests usually have an existing method to derive or compute the $p$-value based on the null distribution. Here we establish the method to estimate a $p$-value for a visual test.

Within the context of visual inference, with $K$ independent observers, the visual $p$-value can be seen as the probability of having as many or more participants detect the data plot than the observed result.

The approach used in @majumder2013validation is as follows. Define $X_j = \{0,1\}$ to be a Bernoulli random variable measuring whether participant $j$ detected the data plot, and $X = \sum_{j=1}^{K}X_j$ be the total number of observers who detected the data plot. Then, by imposing a relatively strong assumption that all $K$ evaluations are fully independent, under $H_0$ $X \sim \mathrm{Binom}_{K,1/m}$. Therefore, the $p$-value of a lineup of size $m$ evaluated by $K$ observer is estimated with $P(X \geq x) = 1 - F(x) + f(x)$, where $F(.)$ is the binomial cumulative distribution function, $f(.)$ is the binomial probability mass function and $x$ is the realization of number of observers choosing the data plot.

As pointed out by @vanderplas2021statistical, this basic binomial model is deficient. It does not take into account the possible dependencies in the visual test due to repeated evaluations of the same lineup, or account for when participants are offered the option to select one or more "most different" plots, or none, from a lineup. They suggest three common lineup scenarios: (1) $K$ different lineups are shown to $K$ participants, (2) $K$ lineups with different null plots but the same data plot are shown to $K$ participants, and (3) the same lineup is shown to $K$ participants. Scenario 3 is the most feasible to apply, but has the most dependencies to accommodate for the $p$-value calculation. For Scenario 3, VanderPlas et al propose modelling the probability of plot $i$ being selected from a lineup as $\theta_i$, where $\theta_i \sim \text{Dirichlet}(\alpha)$ for $i=1,...,m$ and $\alpha > 0$. The number of times plot $i$ being selected in $K$ evaluations is denoted as $c_i$. In case participant $j$ makes multiple selections, $1/s_j$ will be added to $c_i$ instead of one, where $s_j$ is the number of plots participant $j$ selected for $j=1,...K$. This ensures $\sum_{i}c_i=K$. Since we are only interested in the selections of the data plot $i$, the marginal model can be simplified to a beta-binomial model and thus the visual $p$-value is given as

```{=tex}
\begin{equation} \label{eq:pvalue-beta-binomial}
P(C \geq c_i) = \sum_{x=c_i}^{K}{K \choose x}\frac{B(x + \alpha, K - x + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+
\end{equation}
```

\noindent where $B(.)$ is the beta function defined as

```{=tex}
\begin{equation} \label{eq:betafunction}
B(a, b) = \int_{0}^{1}t^{\alpha - 1}(1-t)^{b-1}dt,\quad \text{where}\quad a,b>0.
\end{equation}
```

\noindent We extend the equation to non-negative real number $c_i$ by applying a linear approximation

```{=tex}
\begin{equation} \label{eq:pvalue-beta-binomial-approx}
P(C \geq c_i) = P(C \geq \lceil c_i \rceil) + (\lceil c_i \rceil - c_i) P(C = \lfloor c_i \rfloor), \quad \text{for}\quad c_i \in \mathbb{R}_0^+,
\end{equation}
```

\noindent where $P(C \geq \lceil c_i \rceil)$ is calculated using Equation \ref{eq:pvalue-beta-binomial} and $P(C = \lfloor c_i \rfloor)$ is calculated by

\begin{equation} \label{eq:pmf-beta-binomial}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

The parameter $\alpha$ used in Equation \ref{eq:pvalue-beta-binomial} and \ref{eq:pmf-beta-binomial} is usually unknown and will need to be estimated from the survey data. An interpretation of $\alpha$ is that when it is low only a few plots are attractive to the observers and tend to be selected, and when high, most plots are equally likely to be chosen. VanderPlas et al define $c$-interesting plot to be if $c$ or more participants select the plot as the most different. The expected number of plots selected at least $c$ times, $\mathrm{E}[Z_c]$, is then calculated as

```{=tex}
\begin{equation} \label{eq:c-interesting-expectation}
\mathrm{E}[Z_c(\alpha)] = \frac{m}{B(\alpha, (m-1)\alpha)}\sum_{\lceil c \rceil}^{K}{K \choose x} B(x + \alpha, K - x + (m-1)\alpha).\end{equation}
```

With Equation \ref{eq:c-interesting-expectation}, $\alpha$ can be estimated using maximum likelihood estimation. Precise estimation of $\alpha$, is aided by evaluation of Rorschach lineups, where all plots are null plots. In a Rorschach, in theory all plots should be equally likely, but in practice some (irrelevant) visual elements may be more eye-catching than others. This is what $alpha$ captures, the capacity for extraneous features to distract the observer for a particular type of plot display.

@@

The power of a visual test on the other hand, may additionally depend on the ability of the particular participant, as the skill of each individual may affect the number of observers who identify the data plot from the lineup [@majumder2013validation]. To address this issue, @majumder2013validation models the probability of participant $j$ correctly picking the data plot from lineup $l$ using a mixed-effect logistic regression, with participants treated as random effect. Then, the estimated power of a visual test evaluated by a single participant is the predicted value obtained from the mixed effects model. However, this mixed effects model does not work with scenario where participants are asked to select one or more most different plots. In this scenario, having the probability of a participant $j$ correctly picking the data plot from a lineup $l$ is insufficient to determine the power of a visual test because it does not provide information about the number of selections made by the participant for the calculation of the $p$-value (See Equation \ref{eq:pvalue-beta-binomial-approx}). Therefore, we directly estimate the probability of a lineup being rejected by assuming that individual skill has negligible effect on the variation of the power. This assumption essentially averages out the subject ability and helps to simplify the model structure, thereby obviating a costly large-scale experiment to estimate complex covariance matrices. The same model given in Equation \ref{eq:logistic-regression-1-1} is applied to model the power of a visual test.

@@

Our experiment was conducted over three data collection periods to investigate the difference between conventional hypothesis testing and visual inference in the application of linear regression diagnostics. Two types of departures, non-linearity and heteroskedasticity, were collected during data collection periods I and II. The data collection period III was designed primarily to measure human responses to null lineups so that the parameter $\alpha$ in Equation \ref{eq:pvalue-beta-binomial} can be estimated. Additional lineups for both non-linearity and heteroskedasticity, using uniform fitted value distributions, were included for additional data, and to avoid participant frustration of too many difficult tasks. Overall, we collected `r vi_survey  %>% filter(!attention_check) %>% nrow()` evaluations on `r vi_survey  %>% filter(!attention_check) %>% pull(unique_lineup_id) %>% unique() %>% length()` unique lineups performed by `r vi_survey  %>% filter(!attention_check) %>% count(exp, set) %>% nrow()` participants throughout three data collection periods.

@@

Three different sample sizes were used in our experiment: $n = 50, 100, 300$. Figure \ref{fig:different-n} shows the non-linear "S" shape for different sample sizes. We expect signal strength to decline in the simulated data plots with smaller $n$. A sample size of 300 is typically enough for structure to be visible in a scatterplot reliably. Beyond 300, structure can be obscured by over-plotting, and a different type of display that included transparency, binning or density mapping, might be needed.

@@

One of the notable limitations of the lineup protocol lies in its reliance on human judgements. In this context, the effectiveness of the lineup protocol is closely connected with the perceptive ability and subjective interpretation of individual evaluators. This reliance on human judgment introduces an element of variability and subjectivity that can potentially impact the robustness of the results obtained. To ensure reliable outcomes, it's often necessary to include a diverse group of participants capable of offering unbiased judgements. This requirement has led researchers, including those in our study, to turn to crowd-sourcing platforms that offer access to a random selection of participants, which can help mitigate potential bias.

It is worth noting that despite the need for participants with unbiased perspectives, the lineup protocol can be implemented at a relatively low cost. Leveraging tools like the R package `nullabor`, the process of constructing and evaluating lineups becomes accessible and straightforward. This accessibility empowers a wider range of researchers, even those without extensive statistical backgrounds, to employ the lineup protocol in their studies. Moreover, the cost-effectiveness of lineup-based analyses makes them particularly appealing for integration into statistical education within classroom settings (cite adam loy paper here). The practicality of lineup demonstrations can facilitate a deeper understanding of the concept of randomness and statistical testing among students.

However, it's important to understand that if the goal is to attain results similar to those of formal hypothesis testing in rigor and reliability, using crowd-sourcing becomes necessary. The expense associated with crowd-sourcing is relatively reasonable when compared to the costs of other laboratory devices, making it a sensible investment for research that requires clear and valuable findings. Importantly, this higher level of scrutiny is typically justified only when the data plot can not be confidently identified and requires a thorough investigation.

Additionally, there is the exciting possibility of utilizing computer vision technology to automate the analysis of lineup of residual plots. This innovation has the potential to significantly reduce the expenses associated with recruiting human judges. By incorporating computer vision models, researchers can simplify the evaluation process, making it more efficient and cost-effective.

The scope of this study is to highlight practical value of residual plots in model diagnostics, drawing comparisons with conventional tests. Our primary focus is on the use of residual-fitted plots in a controlled model environment for the purpose of a human subject experiment. It's important to note that there are other intriguing aspects of residual diagnostics that fall beyond the boundaries of this paper. For example, as mentioned in Introduction, residual plots can be used to explore interesting data characteristics while conducting model diagnostics. Furthermore, some model violations, such as serial correlation, necessitate the use of alternative residual plots like residual against row number plots. Nevertheless, we believe that the insights derived from this study have the potential to apply to other residual diagnostics setting, because the fundamental principles and statistical concepts underlying the lineup protocol and residual plots remain consistent across different contexts. In real-world situations, researchers often rely on visual assessment to make informed decisions about model validity. The lineup protocol's capability to identify deviations from randomness serves as a valuable aid in this scenario.

@@

Regression analysis experts suggest that residual plots are indispensable methods for assessing model fit. We conducted a perceptual experiment using visual inference to assess this oft-repeated advice. The experiment tested two primary departures from good residuals: non-linearity and heteroskedasticity.

We found that conventional residual-based statistical tests are more sensitive to weak departures from model assumptions than visual tests. That is, a conventional test concludes there are problems with the model fit almost twice as often as a human. Conventional tests often reject when departures in the form of non-linearity and heteroskedasticity are not visibly different from null residual plots.

While it might be argued that the conventional tests are correctly detecting small but real effects, this can also be seen as the conventional tests are rejecting unnecessarily. Many of these rejections happen even when downstream analysis and results would not be significantly affected by the small departures from a good fit. The results from human evaluations provide a more practical solution, which reinforces the statements from regression experts that residual plots are an indispensable method for model diagnostics.

It is important to note that residual plots need to be delivered as a lineup, embedded in a field of null plots. A residual plot may contain many visual features, but some are caused by the characteristics of the predictors and the randomness of the error, not by the violation of the model assumptions. These irrelevant visual features have a chance to be filtered out by participants with a comparison to null plots, resulting in more accurate reading. The lineup enables a careful calibration for reading structure in residual plots.

Human evaluation of residuals is expensive, time-consuming and laborious. This is possibly why residual plot analysis is often not done in practice. However, with the emergence of effective computer vision, our work could lay a foundation for automated residual plot assessment. <!--The findings suggest the strong demand of graphical inspection in regression diagnostics, so developing an automatic visual inference system to evaluate lineups of residual plots is valuable. We plan to build a completed open-source system in an `R` package and provide a web interface such as a website for public to interact with. Details about this system will be discussed in our next paper.-->

>>>>

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the availability of many numerical hypothesis test procedures designed to use residuals to assess problems with a model fit. Here we provide evidence for why this is good advice using data from a visual inference experiment. We show how conventional tests are too sensitive, which means that too often the conclusion would be that the model fit is inadequate. The experiment uses the lineup protocol which puts a residual plot in the context of null plots. This helps generate reliable and consistent reading of residual plots for better model diagnosis. It can also help in an obverse situation where a conventional test would fail to detect a problem with a model due to contaminated data. The lineup protocol also detects a range of departures from good residuals simultaneously.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \doublespacing
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

It is also common to make scatterplots of residuals against the fitted values, and each of the explanatory variables, can be used to scrutinize their relationships. If there are any visually discoverable associations, the model is potentially inadequate or incorrectly specified. \textcolor{red}{It is also possible to discover patterns not directly connected to a linear model assumption from these residual plots, such as the discreteness or skewness of the fitted values, and outliers.} To read residual plots, one looks for noticeable departures from the model such as non-linear pattern or heteroskedasticity. A non-linear pattern would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption. Statistical tests were developed to provide objective assessment, for example, of non-linear patterns [e.g. @ramsey1969tests], and heteroskedasticity  [e.g. @breusch1979simple].

\textcolor{red}{There are also many variations on residual plots that are important to consider making. The normal probability plot is used to assess a normality assumption on the error distribution. Residuals are often plotted against  order of observations to check for collection bias, batch effects or autocorrelation. Partial residual plots assess the importance of an independent variable to a model, conditional on other variables, by plotting partial residuals against this variable. Partial residuals are computed from the difference between the response value and the fitted model excluding term of the independent variable of interest. A partial regression plot assesses the unique effect of an independent variable in a multiple regression model, by plotting residuals from the model with and without the this variable against each other.}

<!--other types of residual plot to extract the available information and identify inappropriate assumptions, like the added variable plot (cite here) for studying the role of an external variable if it enters linearly into a model. Partial residual plots (cite here) have been suggested as computationally convenient substitutes for the added variable plots. For use in diagnostic procedures, several transformations of the ordinary residuals have been suggested to overcome partially some of their shortcomings. Studentized residuals (cite here) have null distributions that are independent of the scale parameters. These residuals are shown to be closely related to a mean shift model for outliers and to the residuals obtained when each case in turn is left out of the data. Alternatively. the residuals can be transformed to have a selected covariance structure: The usual suggestion is to obtain residuals with uncorrelated elements.}-->

These are all topics explained thoroughly in the sizeable literature on diagnostic procedures for linear models, e.g. @draper1998applied, @montgomery1982introduction, @belsley1980regression, @cook1999applied and @cook1982residuals. Interestingly, despite the abundance of conventional tests, ALL of these writings advise that plotting residuals is an essential tool for diagnosing regression model problems:

@@

\noindent The common wisdom of experts is that plotting the residuals is indispensable for diagnosing model fits. The lack of empirical evidence for the ubiquitous advice is *curious*, and is what this article tackles.

\textcolor{red}{However, it needs to be noted that relying purely on subjective assessment of plots can be a source of anxiety. Indeed, correctly judging whether NO pattern exists in a residual plot is a really difficult task. People will almost always see a pattern (see @kahneman2011thinking), so the question that really needs answering is whether any pattern perceived is consistent with randomness, or sampling variability, or noise. @loy2021bringing emphasizes that this is especially difficult to teach to new analysts and students, and advocates to the broader use of the lineup protocol [@bujastatistical2009].}

\textcolor{red}{The lineup protocol places a data plot in a field of null plots, allowing for a comparison of patterns due purely by chance to what is perceived in the data plot. For residual analysis this is especially helpful for gauging whether there is NO pattern. In its strict use, one would insist that the data plot is not seen before seeing the lineup, so that the observer doesn't know which is the true plot. When used this way, it provides an objective test for data plots. @majumdervalidation2013 validated that results from lineups assessed by human observers performed similarly to conventional tests. One wouldn't use a lineup when a conventional test exists and is adequate because it is more manually expensive to conduct. However, where no adequate conventional test exists, it is invaluable, as shown by @loy2013diagnostic. Here we use the lineup as a vehicle to rigorously explore why experts advise that residual plots are indispensable despite the prevalence of numerical tests.}

@@

In hypothesis testing, a $p$-value is defined as the probability of observing test results at least as extreme as the observed result assuming $H_0$ is true. Conventional hypothesis tests usually have an existing method to derive or compute the $p$-value based on the null distribution. The method to estimate a $p$-value for a visual test essentially follows the process detailed by @vanderplas2021statistical. Details are given in Appendix A.

@@

The power of a visual test on the other hand, may additionally depend on the ability of the particular participant, as the skill of each individual may affect the number of observers who identify the data plot from the lineup [@majumder2013validation]. To address this issue, @majumder2013validation models the probability of participant $j$ correctly picking the data plot from lineup $l$ using a mixed-effect logistic regression, with participants treated as random effect. Then, the estimated power of a visual test evaluated by a single participant is the predicted value obtained from the mixed effects model. However, this mixed effects model does not work with scenario where participants are asked to select one or more most different plots. In this scenario, having the probability of a participant $j$ correctly picking the data plot from a lineup $l$ is insufficient to determine the power of a visual test because it does not provide information about the number of selections made by the participant for the calculation of the $p$-value. Therefore, we directly estimate the probability of a lineup being rejected by assuming that individual skill has negligible effect on the variation of the power. This assumption essentially averages out the subject ability and helps to simplify the model structure, thereby obviating a costly large-scale experiment to estimate complex covariance matrices. The same model given in Equation \ref{eq:logistic-regression-1-1} is applied to model the power of a visual test.

@@

Our experiment was conducted over three data collection periods to investigate the difference between conventional hypothesis testing and visual inference in the application of linear regression diagnostics. Two types of departures, non-linearity and heteroskedasticity, were collected during data collection periods I and II. The data collection period III was designed primarily to measure human responses to null lineups so that the visual $p$-values can be estimated. Additional lineups for both non-linearity and heteroskedasticity, using uniform fitted value distributions, were included for additional data, and to avoid participant frustration of too many difficult tasks. Overall, we collected `r vi_survey  %>% filter(!attention_check) %>% nrow()` evaluations on `r vi_survey  %>% filter(!attention_check) %>% pull(unique_lineup_id) %>% unique() %>% length()` unique lineups performed by `r vi_survey  %>% filter(!attention_check) %>% count(exp, set) %>% nrow()` participants throughout three data collection periods.

@@

Three different sample sizes were used in our experiment: $n = 50, 100, 300$. Figure \ref{fig:different-n} shows the non-linear "S" shape for different sample sizes. We expect signal strength to decline in the simulated data plots with smaller $n$. We chose 300 as the upper limit, because it is typically enough for structure to be visible in a scatterplot reliably. \textcolor{red}{Beyond 300, the scatterplot should probably be used with transparency or replaced with a density or binned plot. Scatterplots suffer from over-plotting.}

@@

One of the primary limitations of the lineup protocol lies in its reliance on human judgements. In this context, the effectiveness of a single lineup evaluation is closely connected with the perceptual ability and visual skills of the individual. When results from multiple individuals are combined the outcome is encouragingly robust. For simple plots and strong patterns just a few individuals are needed to arrive at a clear answer, but more individuals will be needed when the plot design is complex, or the signal strength is weak. <!--This reliance on human judgment introduces an element of variability and subjectivity that can potentially impact the robustness of the results obtained. To ensure reliable outcomes, it's often necessary to include a diverse group of participants capable of offering unbiased judgements. This requirement has led researchers, including those in our study, to turn to crowd-sourcing platforms that offer access to a random selection of participants, which can help mitigate potential bias.-->

Using a lineup protocol removes subjectiveness in interpreting patterns in plots. A plot is compared with draws from a null model, in much the same way as a test statistic is compared to its sampling distribution. It is important to define the data plot using the language provided by a grammar of graphics (e.g. using @ggplot2) because this provides the functional mapping of abstract variables to plot elements. It is also important to remove plot elements that might introduce bias, such as axis labels, text and legends, or to make them generic.

The lineup protocol can be used cheaply and informally with the R package `nullabor`. There is evidence that it is being used fairly broadly, based on software download rates and citations of the original papers. For residual plot analysis we recommend that the lineup be the default first plot so that the data plot is only seen in the context of null plots. When a rigorous test is needed, we recommend using a crowd-sourcing service, as done in gene expression experiment described in @RNAseq2013. While it takes extra effort it is not difficult today, and costs are tiny compared to the overall costs of conducting a scientific experiment. We do also expect that at some point a computer vision model can be developed to take over the task of employing people to evaluate residual plots.

<!--the process of constructing and evaluating lineups becomes accessible and straightforward. This accessibility empowers a wider range of researchers, even those without extensive statistical backgrounds, to employ the lineup protocol in their studies. Moreover, the cost-effectiveness of lineup-based analyses makes them particularly appealing for integration into statistical education within classroom settings (cite adam loy paper here). The practicality of lineup demonstrations can facilitate a deeper understanding of the concept of randomness and statistical testing among students.-->

<!--However, it's important to understand that if the goal is to attain results similar to those of formal hypothesis testing in rigor and reliability, using crowd-sourcing becomes necessary. The expense associated with crowd-sourcing is relatively reasonable when compared to the costs of other laboratory devices, making it a sensible investment for research that requires clear and valuable findings. Importantly, this higher level of scrutiny is typically justified only when the data plot can not be confidently identified and requires a thorough investigation.-->

For this study, simulated data was used to provide a precisely controlled environment within which to compare results from conventional testing to those from visual testing. We also explored only the most commonly used, the residual vs fitted value plots. However, we expect the behavior of the conventional test and the visual test to be similar when observed residuals are diagnosed with this type of plot or other residual plots. The conventional tests will be more sensitive to small departures from the null. They will also fail to detect departures when delivered residuals that have some contamination, like outliers or anomalies, as is often encountered when working with data. The lineup approach is well-suited for generally interpreting data plots, and also detecting unexpected patterns not related to the model. This is supported by earlier research (e.g. @wickham2010, @roy2015using, @loy2015you, @VanderPlas2015, @loy2016variations).

<!-- Additionally, there is the exciting possibility of utilizing computer vision technology to automate the analysis of lineup of residual plots. This innovation has the potential to significantly reduce the expenses associated with recruiting human judges. By incorporating computer vision models, researchers can simplify the evaluation process, making it more efficient and cost-effective.

The scope of this study is to highlight practical value of residual plots in model diagnostics, drawing comparisons with conventional tests. Our primary focus is on the use of residual-fitted plots in a controlled model environment for the purpose of a human subject experiment. It's important to note that there are other intriguing aspects of residual diagnostics that fall beyond the boundaries of this paper. For example, as mentioned in Introduction, residual plots can be used to explore interesting data characteristics while conducting model diagnostics. Furthermore, some model violations, such as serial correlation, necessitate the use of alternative residual plots like residual against row number plots. Nevertheless, we believe that the insights derived from this study have the potential to apply to other residual diagnostics setting, because the fundamental principles and statistical concepts underlying the lineup protocol and residual plots remain consistent across different contexts. In real-world situations, researchers often rely on visual assessment to make informed decisions about model validity. The lineup protocol's capability to identify deviations from randomness serves as a valuable aid in this scenario.-->

@@

This paper has described experimental evidence providing support for the advice of regression analysis experts *that residual plots are indispensable methods for assessing model fit*, using the formal framework of the lineup protocol. We conducted a perceptual experiment on scatterplots of residuals vs fitted values, with two primary departures from good residuals: non-linearity and heteroskedasticity. We found that conventional residual-based statistical tests are more sensitive to weak departures from model assumptions than visual tests. That is, a conventional test concludes there are problems with the model fit almost twice as often as a human. Conventional tests often reject when departures in the form of non-linearity and heteroskedasticity are not visibly different from null residual plots.

While it might be argued that the conventional tests are correctly detecting small but real effects, this can also be seen as the conventional tests are rejecting unnecessarily. Many of these rejections happen even when downstream analysis and results would not be significantly affected by the small departures from a good fit. The results from human evaluations provide a more practical solution, which reinforces the statements from regression experts that residual plots are an indispensable method for model diagnostics. Further work would be needed to *quantify* how much departure from good residuals is too much.

It is important to emphasize that this work also supports a change in common practice, which is to deliver residual plots as a lineup, embedded in a field of null plots, rather than be viewed out of context. A residual plot may contain many visual features, but some are caused by the characteristics of the predictors and the randomness of the error, not by the violation of the model assumptions. These irrelevant visual features have a chance to be filtered out by participants with a comparison to null plots, resulting in more accurate reading. The lineup enables a careful calibration for reading structure in residual plots, and also provides the potential to discover interesting and important features in the data not directly connected to linear model assumptions.

Human evaluation of residuals is expensive, time-consuming and laborious. This is possibly why residual plot analysis is often not done in practice. However, with the emergence of effective computer vision, it is hoped this work helps to lay the foundation for automated residual plot assessment. <!--The findings suggest the strong demand of graphical inspection in regression diagnostics, so developing an automatic visual inference system to evaluate lineups of residual plots is valuable. We plan to build a completed open-source system in an `R` package and provide a web interface such as a website for public to interact with. Details about this system will be discussed in our next paper.-->
